---
title: "Assessing model fit"
format: html
author: Andrew MacDonald and Catherine Čapkun-Huot
theme: minty
bibliography: references.bib
comments: 
  hypothesis: true
code-fold: true
---

In this paper we present a new model for Flight Initiation distance (FID). This model captures the process of habituation in three parameters (see Appendix 3). However, this model contrasts with the log-linear models common in the literature, for example as used by Allan et al. [-@allan2020]

In order to asses fit of our model and compare it to a log-linear model, we use posterior predictive checks. Posterior predictive checks are key components of a Bayesian workflow [@gabry2019]. In a posterior predictive check, we simulate observations using the model and compare these to the original data. A model that fits the data well should generate predictions that closely resemble that data.

```{r message=FALSE}
library(brms)
library(tidyverse)
# Load FID data - data_FID_analyses.csv -, each row is a FID trial
data_FID <- read.csv(here::here("Data/data_FID_analyses.csv"), 
                     header = TRUE, 
                     sep = ",")

# Data structure
data_FID <- tibble(ID      = as.factor(data_FID$ID),
                   num_obs = as.numeric(data_FID$obs_number),
                   FID     = as.numeric(data_FID$FID + 0.01), # gamma distribution needs values > 0 
                   risk    = as.factor(data_FID$Approach_speed))
```

```{r}
# Formula
form_fit1 <- bf(FID    ~ inv_logit(logitM) * 1000 * (1 - inv_logit(logitp)*num_obs/(exp(logd) + num_obs)),
                logitM ~ 1 + risk + (1 |t| ID),
                logitp ~ 1 + risk + (1 |t| ID),
                logd   ~ 1 + risk + (1 |t| ID),
                nl     = TRUE,
                family = Gamma(link = "identity"))

# Assign priors
priors <- c(prior(lkj(2), class = "cor"),
            prior(exponential(2),   class = "sd", nlpar = "logd"),
            prior(exponential(2),   class = "sd", nlpar = "logitM"),
            prior(exponential(2),   class = "sd", nlpar = "logitp"),
            prior(normal(1.5, .5),  class = "b",  nlpar = "logd"),
            prior(normal(.5,.5),    class = "b",  nlpar = "logitM"),
            prior(normal(1, .2),   class = "b",  nlpar = "logitp"),
            prior(gamma(6.25, .25), class = "shape"))

# Run model 1
fit1 <- brm(form_fit1,
            data        = data_FID,
            prior       = priors,
            seed        = 1234,
            adapt_delta = 0.95,
            core        = 4,
            iter        = 2000,
            backend     = "cmdstanr",
            file        = here::here("Appendix_1/fit1.rds"),
            file_refit  = "on_change")
```

```{r}
## contrast this with a log model
form_loglin <- bf(log(FID) ~ 1 + num_obs + risk + risk:num_obs + (1 + num_obs | ID),
                family = gaussian())


prior_loglin <- c(
  prior(normal(0, .5), class = "b"),
  prior(normal(0, .5), class = "Intercept"),
  prior(lkj(2), class = "cor"),
  prior(exponential(2), class = "sd"),
  prior(exponential(1), class = "sigma")
)

fit_loglin <-  brm(form_loglin,
                   data        = data_FID,
                   prior       = prior_loglin,
                   seed        = 1234,
                   adapt_delta = 0.95,
                   core        = 4,
                   iter        = 2000,
                   backend     = "cmdstanr",
                   file        = here::here("Appendix_1/fit_loglin"), 
                   file_refit  = "on_change")
```

```{r}
#| label: fig-comp
#| fig-cap: comparison of our model with a log-linear model, which is a common alternative for analyzing FID data. Dark blue lines indicate the observed data, light blue lines are simulated observations from the posterior predictive distribution.
#| fig-subcap: 
#|   - Linear model on log-transformed FID data.
#|   - Our proposed Model 1, fit to the same data.
#| layout-ncol: 2

# summary(fit1)


p1 <- brms::pp_check(fit_loglin, "dens_overlay") + 
  labs(title = "Log-transformed linear model")

p2 <- brms::pp_check(fit1, "dens_overlay") + 
  labs(title = "Model 1 from Čapkun-Huot et al.")

p1
p2

```

Figure 1 shows the fit between our model and the log-linear model which is common in studies of FID. The log-linear approach represents the average well, but overestimates the variance in FID. In contrast, our model captures variance well, though it underestimates slightly the mode of the dataset.
